---
title: "Lab 01 — Introduction to R and the Creativity Study"
author: "Ashish Pant"
date: "2026-05-21"
format: html
---

## 1. Rocks

1.  The one that I believe best represents the average weight of the population is 13.
2.  5 rocks that I believe best represents the average weight of the population are 13, 7, 83, 81 and 62.
3.  5 rocks that best represent variability are 67, 69, 13, 95 amd 84

## 2. Introduction to Computing

### Basics

Load necessary packages

```{r}
library(dplyr)
library(ggplot2)
```

Read CSV

```{r}
creativity <- read.csv("../data/creativity.csv")
names(creativity)
```

Print the first 6 (default) observations

```{r}
head(creativity)
```

Quick six-number summary of numerical column score

```{r}
summary(creativity$score)
```

Summarize all variables

```{r}
summarize(creativity)
```

### Using the "pipe" operator \|\>

```{r}
creativity |>
    head()
```

Select columns from a data frame

```{r}
creativity |>
    select(score) |>
    head()
```

Return six-number summary on the score column

```{r}
creativity |>
    pull(score) |> # pull extracts a vector of the column values
    summary()
```

Return the mean of the score column

```{r}
creativity |>
    pull(score) |>
    mean()
```

Return mean using summarize:

```{r}
creativity |>
    summarize(
        meanScore = mean(score)
    )
```

Summarize many expressions

```{r}
creativity |>
    summarize(
        meanScore = mean(score),
        medianScore = median(score),
        sdScore = sd(score)
    )
```

Calculate summaries by treatment group

```{r}
creativity |>
    group_by(treatment) |>
    summarize(
        meanScore = mean(score),
        medianScore = median(score),
        sdScore = sd(score)
    )
```

### Plotting Data

Create a dot plot

- Center all dots on x-axis value 1

- y is the score variable

- binaxis is the axis to bin along, using y here

- stackdir specifies direction in which dots are stacked

```{r}
ggplot(data = creativity, mapping = aes(x = 1, y = score)) +
    geom_dotplot(binwidth = 1, binaxis = "y", stackdir = "center")
```

Create dot plots by treatment group

- Use the treatment column in the x-axis

```{r}
ggplot(data = creativity, mapping = aes(x = treatment, y = score)) +
    geom_dotplot(binwidth = 1, binaxis = "y", stackdir = "center")
```

Avoid redundancy by building a template and adding different geom layers to the template

- Create a template

```{r}
creativity_setup1 <- ggplot(data = creativity, mapping = aes(x = score))
```

- Add geom layer - histogram

```{r}
creativity_setup1 + geom_histogram()
```

- Add geom layer box_plot

```{r}
creativity_setup1 + geom_boxplot()
```

- Flip the box plot x and y axes

```{r}
creativity_setup1 + 
    coord_flip() +
    geom_boxplot()
```

Work with treatment group

```{r}
creativity_setup2 <- ggplot(data = creativity, mapping = aes(x = treatment, y = score))
```

- Draw box plot per treatment group

```{r}
creativity_setup2 + geom_boxplot()
```

- Change theme to classic theme and redraw the same box plot

```{r}
creativity_setup2 + 
    geom_boxplot() +
    theme_classic()
```

Specify different colors to visually separate groups

- Create the template

```{r}
creativity_setup3 <- ggplot(
    data = creativity, 
    mapping = aes(
        x = score,
        fill = treatment,
        color = treatment
    )
)
```

- Draw histogram

```{r}
creativity_setup3 + 
    geom_histogram(position = "dodge")
```

- Change color scale

```{r}
creativity_setup3 +
    scale_color_grey() + 
    scale_fill_grey() +
    geom_histogram(position = "dodge")
```

Save the last plot

```{r}
ggsave("figure.png")
```

Save a plot for HW1 submission:

```{r}
creativity_setup2 + 
    geom_boxplot() +
    theme_classic()

ggsave("creativity_boxplot.png", width = 6, height = 4, dpi = 300)
```

### Self-Assessment

Read tomato.csv

```{r}
tomato_df <- read.csv("../data/tomato.csv")
names(tomato_df)
```

View the first 6 records

```{r}
head(tomato_df)
```

Draw side-by-side box plots of each yield for each of the two fertilizers

```{r}
ggplot(data = tomato_df, mapping = aes(x = treatment, y = yield)) +
    geom_boxplot() +
    theme_classic()
```

Estimate the difference in mean yield between the two fertilizers (a - b). Include units.

```{r}
mean_a = tomato_df |> filter(treatment == "a") |> pull(yield) |> mean()
mean_b = tomato_df |> filter(treatment == "b") |> pull(yield) |> mean()
paste("Mean difference: ", round(mean_a - mean_b, 2), "pounds")
```

Estimate the difference between in median yield the two fertilizers (a - b). Include units.

```{r}
median_a = tomato_df |> filter(treatment == "a") |> pull(yield) |> median()
median_b = tomato_df |> filter(treatment == "b") |> pull(yield) |> median()
paste("Median difference: ", round(median_a - median_b, 2), "pounds")
```