---
title: "Lab 02 — Standard Errors and Randomization Tests"
author: "Ashish Pant"
date: "2026-05-26"
format: html
---

### Load libraries

Load tidyverse library

```{r}
library(tidyverse)
library(here) # resolves files from the project root regardless of where qmd lives
```

### Writing functions in R

Load data (user header=TRUE)

```{r}
creativity <- read.csv(here("data/creativity.csv"), header = TRUE)
head(creativity)
```

Define function for computing the standard error

```{r}
se <- function(x) {
    s <- sd(x, na.rm = TRUE)
    n <- sum(!is.na(x))
    s / sqrt(n)
}
```

Calculate standard error for all 47 observation scores (just the score column)

```{r}
se(creativity$score)
```

Using tidyverse, summarize the standard error (a different way to calculate)

```{r}
creativity |>
    summarize(
        se = se(score)
    )
```

Group by treatment and compute the standard error by treatment group

```{r}
creativity |>
    group_by(treatment) |>
    summarize(
        se = se(score)
    )
```

### Randomization Tests

First set seed for reproducibility

```{r}
set.seed(5101)
```

Calculate the observed difference in mean score (observed means as documented in creativity.csv)

```{r}
observed_diff <- creativity |>
    group_by(treatment) |>
    summarize(mean_score = mean(score)) |>
    summarize(diff = diff(mean_score)) |>
    pull(diff)

observed_diff
```

Run randomization (basically, simulate what else, other than what was recorded in the csv, could have been observed at random)

```{r}
nsim <- 10000
simulated_diffs <- replicate(nsim, {
    # Everthing in {} is evaluated as if it was a function (this is like a lambda)
    creativity |>
        # generate a random treatment value
        mutate(shuffled = sample(treatment)) |>
        # group by the random treatment value
        group_by(shuffled) |>
        # Calculate the mean and diff like before
        summarize(mean_score = mean(score)) |>
        summarize(diff = diff(mean_score)) |>
        # pull the difference
        pull(diff)
})
head(simulated_diffs)
```

Calculate the p-value

```{r}
p_value <- (1 + sum(abs(simulated_diffs) >= abs(observed_diff))) / (nsim + 1)
p_value
```

Explanation of p-value calculation: The p-value is the number of observations with a value equal to or more extreme (\>=) than the observed value divided by the number of observations. This would translate to:

sum(abs(simulated_diffs) \>= abs(simulated_diff)) / nsim

But, this results in a 0 p-value if by chance the observed_diff was an extreme value itself and none of the simulated values were equal or larger resulting in 0 / nsim = 0, which is an impossible p-value.

By adding 1 to both the numerator and denominator, essentially, we are including the observed_diff in the calculation. Since observed_diff is equal to itself, it is valid to include it as a count in the numerator and increment the denominator by 1 as if we performed 10001 simulations instead of 10000.

### Self Assessment

Read in the tomato data, tomato.csv

```{r}
tomato <- read.csv(here("data/tomato.csv"), header = TRUE)
head(tomato)
```

#### (a) Draw side-by-side box plots of the yield for each fertilizer

```{r}
ggplot(data = tomato, mapping=aes(x = treatment, y = yield)) +
    geom_boxplot()
```

#### (b) Estimate the difference in mean yield between the two fertilizers (a-b)

```{r}
observed_diff = tomato |>
    group_by(treatment) |>
    summarize(mean_yield = mean(yield)) |>
    summarize(diff = diff(mean_yield)) |>
    pull(diff)
observed_diff
```

#### (c) Estimate the difference in median yield

```{r}
tomato |>
    group_by(treatment) |>
    summarize(median_yield = median(yield)) |>
    summarize(diff = diff(median_yield)) |>
    pull(diff)

```

#### (d) Use a randomization test to test the null hypothesis of no effect of the fertilizer on mean yield. Report the two-sided p-value

Run randomization test and collect simulated diffs

```{r}
# Set seed for reproducibility
set.seed(5101)

# We have already calculated the observed diff above
# Run randomization test

nsim <- 10000
simulated_diffs <- replicate(nsim, {
    tomato |>
        mutate(shuffled = sample(treatment)) |>
        group_by(shuffled) |>
        summarize(mean_yield = mean(yield)) |>
        summarize(diff = diff(mean_yield)) |>
        pull(diff)
})
head(simulated_diffs)
```

Calculate p-value

```{r}
p_value = (1 + sum(abs(simulated_diffs) >= abs(observed_diff))) / (nsim + 1)
p_value
```

Comments: Treatment a (the proposed better fertilizer) produced a mean yield 6.53 pounds higher than treatment b (p = 0.037). This provides moderate evidence against the null hypothesis of no fertilizer effect — a difference this large would occur by chance alone only about 3.7% of the time under random assignment. Since the 7 plants per group were randomly assigned to treatments, a causal conclusion is justified: the better fertilizer caused higher yield in this experiment. However, since the 14 plants were not randomly selected from any broader population, we cannot generalize beyond these particular plants.